\myTitle

### 2025-03-17 tex/body-text-of-paper.tex
Pushing the revisions further along

@@

We present a new reference baseline that includes 1,636 Chinook salmon individuals from 17 collections that were genotyped at 204 loci distributed throughout the genome.
This baseline is highly effective for GSI within California and provides ample power for PBT, thus enabling a highly effective and efficient integrated GSI/PBT monitoring and evaluation program 
\citep{garza2007large,beacham2021parentage}.
We describe how we identified and compiled the markers, including several new markers
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

We compiled samples of fish for the reference baselines from 13 locations within California, and one
within Oregon. This set of populations and stocks includes all of the previously described lineages of Chinook salmon in the southern portion of the range and most of the populations that are known to have significant genetic differentiation in this region.
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
three rivers of the California coast (Russian, Eel, and Smith).
Of the 1594 total fish samples from 17 different collections, 724 fish from 10 collections were of the fall run ecotype,
498 fish from 5 collections were spring run, and 261 and 111 fish were of the winter run and late-fall run ecotypes, 
respectively,  each represented by a single collection.
(Figure~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run.
Names of each location are followed by the location codes (Table~\ref{tab:samples}).
Abbreviations used in location names as follows:
R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

Collections in each location were separated into fish from the different adult migration timing ecotypes according
to a variety of criteria.  Winter-run Chinook salmon are propagated at the Livingston
Stone National Fish Hatchery for supplementation (released as juveniles into the river) and as a captive broodstock;
sampling of this population was performed by hatchery staff during spawning. In Central Valley rivers with both
spring-run and fall-run ecotypes (i.e. Mill, Deer and Butte Creeks), previous research has documented different  (but overlapping) ranges of spawning time (Julian date) for each ecotype \citep{fry1961king,yoshiyama1998historical}.
As such, sampling targeted carcasses found in each of these
rivers at times representative of their run-timing designation, which we were able to further confirm using their genotypes
within the region of strongest association (RoSA) on chromosome 28 \citep{thompson2020complex}.
In the San Joaquin River, in 2016, at the time of sampling, only
fall-run Chinook salmon adults were found, therefore the samples could be categorized.  Run-timing
designation at the Feather River Hatchery is more complicated but it is managed by the hatchery practices in place:
early returning fish in May and April (expressing the spring-run life history) are visibly tagged and returned
to the river. Subsequently in the fall, when hatchery raceways are reopened, only fish with the visible tag are included
as spring-run broodstock, and the rest are included in the fall-run broodstock.
We sampled putative spring-run and fall-run fish 
from these two broodstocks. Similar to broodstock collections in the CCV, samples from spring- and fall-run fish in the Trinity River were selected on the basis of spawn timing at the Trinity River Hatchery with migration-timing ecotype then confirmed by RoSA genotype. Elsewhere, adult samples were taken from live broodstock at Rowdy Creek (Smith River), Iron Gate (Klamath River) and Warm Springs (Russian River) Hatcheries and juveniles were sampled from Blue Creek (Klamath River); all locations where only fall-run fish have been documented.  Spring run adult fish were sampled from the spring-run broodstock at Cole Rivers Hatchery on the Rogue River. Samples from the Eel River were taken from live fish ascending a ladder in the upper basin and were confirmed as fall-run ecotype by RoSA genotype.

@@

The genetic markers in the baseline include newly discovered microhaplotypes, SNP assays
that were translated into amplicons, markers associated with phenotypes, as well
as species- and sex-specific markers.
These genetic markers were compiled from several different
efforts, as detailed in the sections below.

@@

We identified candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). We used double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citep{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.

Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, a {\em de novo} assembly was performed.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.
Gene regions were then selected that (1) contained multiple SNPs within a 100 bp window, and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.

These gene regions were PCR-amplified and sequenced on a MiSeq sequencer using a $2\times 75$-cycle kit.
The sequences from each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignmets were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, as well as those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
In order to type these markers using amplicon sequencing, we followed the methods in
\citet{campbell2015genotyping} and added the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes \citep{garrison2012haplotype} was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

To identify loci with large allele frequency differences between late-fall and fall-run Chinook salmon,
we first conducted a simple case-control genome-wide association study (GWAS) using the low-coverage whole-genome sequencing
data of \citet{thompson2020complex}, mapped to the Otsh\_v1.0 Chinook salmon assembly
(GenBank accession GCA\_002872995.1).  We conducted
two separate association-study comparisons.
The first involved the 16 late-fall-run fish as cases with 16 fall-run fish from the Feather River Hatchery as controls.
The second involved the same 16 late-fall fish as cases, but used 16 fall-run fish from the San
Joaquin River as controls.  The association studies were performed for each chromosome with
ANGSD \citep{pmid21663684,korneliussen_angsd_2014}, using the following command: {\footnotesize\tt angsd -yBin \{ybin\}  -r \{chrom\}
-minMapQ 30 -minQ 20 -minInd 12 -doAsso 1 -GL 1 -doMajorMinor 1 -doMaf 1 -SNP\_pval 1e-6
-out \{prefix\}  -bam \{bamlist\}}, where {\tt \{ybin\} } is the path to a file indicating the cases and
controls,
{\tt \{chrom\} } indicates the chromosome name, {\tt \{prefix\} } is the prefix to use for output files, and
{\tt \{bamlist\} } is the path to the text file holding paths to the binary alignment map (BAM) files
of aligned sequence data, one for each individual, ordered as in {\tt \{ybin\}}.

We surveyed the GWAS $p$-values relative to the commonly used significance threshold for
GWAS of $5\times 10^{-8}$ \citep{chen2021revisiting}.  Only a single site exceeded that threshold in both comparisons and it occurred
atop a peak in association scores on chromosome 34 (see
Results). From this peak we identified candidates for follow-up genotyping by including the 10
SNPs in that peak with the lowest association $p$-values, and supplementing those with additional SNPs that
showed:
\begin{itemize}
\item An estimated absolute allele frequency difference between late-fall and all 32 fall-run fish, $|d| >0.75$,  with an approximate lower confidence interval of $|d|$ greater than 0.5,  or
\item $|d| >0.5$ with a lower confidence interval $>$0.25 and
with either fall-run or late-fall run being nearly fixed for one of the alleles, or
\item An annotation from snpEff \citep{cingolani2012program} of ``High'' or ``Moderate,'' or
\item $|d|>0.48$ and a lower confidence interval of $>$0.25 and a
genomic  coordinate $> 2.2$~Mb.
\end{itemize}
The final condition was implemented in order to gather several SNPs with large allele frequency differences
that were adjacent to, but not directly within, the main peak of association, as we had hoped that those SNPs
might enable us to learn about recombination in the region.

For the candidate markers found by the above process, we designed amplification primers using
Primer3 \citep{koressaar2007enhancements,untergasser2012primer3} and tested amplification in a
subset of Chinook salmon
samples. The primer pairs that amplified sequence that reliably mapped back to the association-peak region were then typed
in a separate collection of 1,638 Sacramento River Chinook salmon sampled from the fish trap below Keswick Dam, including 1,169 winter run, 281 spring
run, and 188 fall/late-fall run. We divided the fish identified
from the 125-locus genetic
stock identification data set of \citet{thompson2020complex}  as fall/late-fall into fall run and late-fall run according
to their sampling date at Keswick:  99 fish arriving before Dec 26 were classified as fall run and 89 fish arriving after Dec 26 were
classified as late-fall run.   The allele frequencies of the target SNPs at the remaining amplicons
were
estimated for each of these four groups: winter, spring, fall, and late-fall, and we retained
amplicons showing large allele frequency differences between late-fall fish and all other runs.

@@

Using the whole genome sequencing data of \citet{thompson2020complex}, we also sought
genomic regions that might provide diagnostic markers for the Sacramento winter-run
Chinook salmon.  This ecotype is highly differentiated from all other Chinook salmon populations
throughout its genome, which complicates the process of finding diverged genome regions using
a simple association study, with only 16 samples, as done to identify the LFAR.  Consequently we pursued a bespoke
analysis to find regions we could expect to have large allele frequency differences between
the winter-run and all other ecotypes of Chinook salmon in California's Central Valley.   This analysis
screened for regions with large allele frequency differences across large blocks of the genome, and is
described in the supplemental text section \ref{sec:wrap-methods}.

@@

We include six amplicons that capture the eight  SNPs within the ``Region of Strongest Assocation'' (RoSA)
between spring-run and fall-run fish, near the ROCK1 and GREB1L genes,  on Chromosome 28 listed in Table~S3 of \citet{thompson2020complex}.
We add three more amplicons in the same RoSA region to the panel, as well as one more amplicon that
allows genotyping of snp670329, which was discovered and described by \citet{thompson2019anthropogenic}.
Including all these markers in the panel allows them to be routinely typed, creating
 a database for understanding the spatial and temporal
distribution of the alleles in this genomic region; however we have omitted these markers for calculating likelihoods for
GSI, since their distribution across populations in the Feather River has been clearly and evidently
influenced by hatchery practices (see Results), it is not necessary to include them for accurate assignment,
and because the 12 SNPs within these three amplicons are typically in profound linkage disequilibrium (LD)
making them inappropriate for use as separate markers in most software for population assignment or GSI, that
typically assumes markers are not in LD within the component populations of the baseline.

@@

The baseline also includes three amplicons with genetic variation
in two genes, VGLL3 and SIX6,
that have been found to be associated with phenotypes related to reproductive maturity in both
Atlantic \citep{barson2015sex} and Pacific \citep{waters2021heterogeneous} salmonids. We include
them both to provide additional variation for population genetic analyses and to
monitor populations for potential phenotypic patterns.

We include an assay from a novel locus that targets the sex-determining region ({\em sdY}) on the Y chromosome in Chinook salmon. We designed this assay to target the SNP described by \citet{bertho2022nonfunctional} that differentiates the functional {\em sdY} and a nonfunctional {\em sdY} pseudogene.  Examination of whole-genome sequence data in multiple populations of California Chinook salmon \citep{thompson2020complex} shows that this
marker characterizes genetic sex in diverse California salmon populations more accurately than previously published assays  \citep{von2015development}.
And, finally, we include marker Ots\_coho001\_05\_32691399 which is fixed for alternate alleles in
coho salmon and Chinook salmon.  This is helpful when samplers have unwittingly
collected coho salmon.

@@

The novel microhaplotype markers and the SNPlicon targets were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both of these
groups of markers we identified their locations within the genome by mapping the
consensus sequences (roughly 100--300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0, GenBank accession GCA\_002872995.1)
and version 2  (Otsh\_v2.0, GenBank accession GCA\_018296145.1) genome assemblies,
using both bwa mem \citep{bwa-mem2009} and the BLAST-like Alignment tool, BLAT \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers so that we can
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

Amplicon sequencing was performed with the Genotyping-in-Thousands-by-sequencing (GT-seq)
method \citep{campbell2015genotyping}. All loci were
multiplexed with primer concentrations ranging from 0.083 uM to 0.5 uM to increase
uniformity of read depths across all loci. Sample normalization was
performed using the SequalPrep Normalization Plate Kit (Applied Biosystems) according
to the manufacturer's instructions. Samples were pooled by plate
and then purified using Agencourt AMPure XP magnetic beads.
Libraries were quantified either by qPCR with the NEBNext Library Quant Kit for Illumina (New
England BioLabs) or with the Qubit dsDNA HS Assay Kit (Thermo Fisher Scientific)
before dilution and pooling for sequencing.
All samples were sequenced on a MiSeq using $2\times 75$-cycle paired-end sequencing protocols.
Primer sequences and concentrations for making the primer pool are in Supplemental Data~1.

@@

Basic data summaries were calculated for each population/collection
in the reference baseline
using simple operations in the R programming language, Version 4.3
\citep{rcore}. We started with a simple tally of the number of alleles
found at each locus.  To provide a summary of the genetic variation and
missing data we further summarized:
\begin {itemize}
\item the average over loci of the number of fish
in which each locus was successfully genotyped, $\bar{N}$;
\item the average number of alleles per locus in each collection with all collections
downsampled to the smallest number of genotyped individuals at each
locus, $\bar{N}_\mathrm{A,ss}$;
\item the fraction of polymorphic
loci in each collection after subsampling to the smallest number of genotyped
fish, $\bar{P}_\mathrm{poly,ss}$;
\item the expected heterozygosity as the average over loci
of one minus the frequencies of homozygotes at each locus expected from the
estimated allele frequencies, $\bar{H}_\mathrm{exp}$; and
\item the observed heterozygosity as the average over
loci of the fraction of heterozygous genotypes, $\bar{H}_\mathrm{obs}$.
\end{itemize}
When subsampling each collection to the minimum number of genotyped
fish across all collections at a locus, we subsampled individuals without
replacement, taking the average of 1,000 different random subsamples.

We calculated \citet{weir1984estimating}'s pairwise $F_\mathrm{ST}$ between
all collections, using the {\footnotesize\tt pp.fst} function from the R
package `hierfstat' \citep{hierfstat}. Finally, we evaluated population
structure in the data by performing unsupervised clustering 
using the program STRUCTURE \citep{pritchard2000inference,falush2003inference}.
For this analysis, we omitted the run-timing associated markers.
We performed 20 replicate runs at each value of the assumed number of
subpopulations, $K$, in $\{2, 3, 4, 5, 6, 7\}$
using default settings of the program.  STRUCTURE output was
a summarized and visualized using CLUMPAK
\citep{kopelman2015clumpak}.

>>>>

A critical ingredient for GSI is the {\em reference data set}, which consists of individuals
sampled from across a range of populations and with each individual genotyped using the same
panel of genetic markers.  When individuals of unknown origin are genotyped
at the same panel of genetic markers, the availability of the reference data set
makes it possible to do inference regarding the population of origin of the
unknown individuals.  Such reference data sets have seen widespread use and
development in salmon management, where they are often called
``genetic baselines," and defined as, ``databases of genotypes from
breeding populations"\citep{seeb2007development}.  Throughout, we will
use ``reference" or ``reference data set'' and ``baseline'' interchangeably; however,
a key point to keep in mind is that as genetic technologies
evolve over time, reference data sets continue to be developed and
refined to exploit the new technologies.

@@

We present a new reference baseline that includes 1,636 Chinook salmon individuals from 17 collections that were genotyped at 204 loci distributed throughout the genome.
This baseline is highly effective for GSI within California and provides ample power for PBT, thus enabling a highly effective and efficient integrated GSI/PBT monitoring and evaluation program 
\citep{garza2007large,beacham2021parentage}.
We describe how we identified and compiled the markers, including several new markers
identified through re-analysis of the whole genome sequencing data of \citet{thompson2020complex},
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

Samples of fish for the reference baseline were compiled from 13 locations within California, and one
within Oregon. This set of populations and stocks includes all of the previously described lineages of Chinook salmon in the southern portion of the range and most of the populations that are known to have significant genetic differentiation in this region.
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
three rivers of the California coast (Russian, Eel, and Smith).
Of the 1594 total fish samples from 17 different populations, 724 fish from 10 collections were of the fall run ecotype,
498 fish from 5 collections were spring run, and 261 and 111 fish were of the winter run and late-fall run ecotypes, 
respectively,  each represented by a single collection.
(Figure~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run.
Names of each location are followed by the location codes (Table~\ref{tab:samples}).
Abbreviations used in location names as follows:
R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

Collections in each location were separated into fish from the different adult migration timing ecotypes according
to a variety of criteria.  Winter-run Chinook salmon are propagated at the Livingston
Stone National Fish Hatchery for supplementation (released as juveniles into the river) and as a captive broodstock;
sampling of this population was performed by hatchery staff during spawning. In Central Valley rivers with both
spring-run and fall-run ecotypes (i.e. Mill, Deer and Butte Creeks), previous research has documented different  (but overlapping) ranges of spawning time (Julian date) for each ecotype \citep{fry1961king,yoshiyama1998historical}.
As such, sampling targeted carcasses found in each of these
rivers at times representative of their run-timing designation, which were further confirmed using their genotypes
within the region of strongest association (RoSA) on chromosome 28 \citep{thompson2020complex}.
In the San Joaquin River, in 2016, at the time of sampling, only
fall-run Chinook salmon adults were found, therefore the samples could be categorized.  Run-timing
designation at the Feather River Hatchery is more complicated but it is managed by the hatchery practices in place:
early returning fish in May and April (expressing the spring-run life history) are visibly tagged and returned
to the river. Subsequently in the fall, when hatchery raceways are reopened, only fish with the visible tag are included
as spring-run broodstock, and the rest are included in the fall-run broodstock.
Putative spring-run and fall-run fish 
from these two broodstocks were sampled. Similar to broodstock collections in the CCV, samples from spring- and fall-run fish in the Trinity River were selected on the basis of spawn timing at the Trinity River Hatchery with migration-timing ecotype confirmed by RoSA genotype. Elsewhere, adult samples were taken from live broodstock at Rowdy Creek (Smith River), Iron Gate (Klamath River) and Warm Springs (Russian River) Hatcheries and juveniles were sampled from Blue Creek (Klamath River); all locations where only fall-run fish have been documented.  Spring run adult fish were sampled from the spring-run broodstock at Cole Rivers Hatchery on the Rogue River. Samples from the Eel River were taken from live fish ascending a ladder in the upper basin and were confirmed as fall-run ecotype by RoSA genotype.

@@

The genetic markers in the baseline include newly discovered microhaplotypes, SNP assays
that were translated into amplicons, markers associated with phenotypes, and
species- and sex-specific markers.
These genetic markers were compiled from several different
efforts, as detailed in the sections below.

@@

Candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations were identified using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). Double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) were used to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citep{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.

Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.
Gene regions were selected that (1) contained multiple SNPs within a 100 bp window, and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.

These gene regions were PCR-amplified and sequenced on a MiSeq sequencer using a $2\times 75$-cycle kit.
The sequences from each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignmets were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
These markers were typed using amplicon sequencing, following the methods in
\citet{campbell2015genotyping} with the addition of the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes \citep{garrison2012haplotype} was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

Loci with large allele frequency differences between late-fall and fall-run Chinook salmon,
were identified with a simple case-control genome-wide association study (GWAS) using the low-coverage whole-genome sequencing
data of \citet{thompson2020complex}, mapped to the Otsh\_v1.0 Chinook salmon assembly
(GenBank accession GCA\_002872995.1).  We conducted
two separate association-study comparisons.
The first involved the 16 late-fall-run fish as cases with 16 fall-run fish from the Feather River Hatchery as controls.
The second involved the same 16 late-fall fish as cases, but used 16 fall-run fish from the San
Joaquin River as controls.  The association studies were performed for each chromosome with
ANGSD \citep{pmid21663684,korneliussen_angsd_2014}, using the following command: {\footnotesize\tt angsd -yBin \{ybin\}  -r \{chrom\}
-minMapQ 30 -minQ 20 -minInd 12 -doAsso 1 -GL 1 -doMajorMinor 1 -doMaf 1 -SNP\_pval 1e-6
-out \{prefix\}  -bam \{bamlist\}}, where {\tt \{ybin\} } is the path to a file indicating the cases and
controls,
{\tt \{chrom\} } indicates the chromosome name, {\tt \{prefix\} } is the prefix to use for output files, and
{\tt \{bamlist\} } is the path to the text file holding paths to the binary alignment map (BAM) files
of aligned sequence data, one for each individual, ordered as in {\tt \{ybin\}}.

The GWAS $p$-values were compared relative to the commonly used significance threshold for
GWAS of $5\times 10^{-8}$ \citep{chen2021revisiting}.  Only a single site exceeded that threshold in both comparisons and it occurred
atop a peak in association scores on chromosome 34 (see
Results). Candidates for follow-up genotyping were identified in the peak by including the 10
SNPs in that peak with the lowest association $p$-values, and supplementing those with additional SNPs that
showed:
\begin{itemize}
\item An estimated absolute allele frequency difference between late-fall and all 32 fall-run fish, $|d| >0.75$,  with an approximate lower confidence interval of $|d|$ greater than 0.5,  or
\item $|d| >0.5$ with a lower confidence interval $>$0.25 and
with either fall-run or late-fall run being nearly fixed for one of the alleles, or
\item An annotation from snpEff \citep{cingolani2012program} of ``High'' or ``Moderate,'' or
\item $|d|>0.48$ and a lower confidence interval of $>$0.25 and a
genomic  coordinate $> 2.2$~Mb.
\end{itemize}
The final condition was implemented in order to gather several SNPs with large allele frequency differences
that were adjacent to, but not directly within, the main peak of association, so as to possibly learn about
recombination in the region.

We designed amplification primers for the candidates using
Primer3 \citep{koressaar2007enhancements,untergasser2012primer3} and tested amplification in a
subset of Chinook salmon
samples. The primer pairs that amplified sequence that reliably mapped back to the association-peak region were typed
in a separate collection of 1,638 Sacramento River Chinook salmon sampled from the fish trap below Keswick Dam, including 1,169 winter run, 281 spring
run, and 188 fall/late-fall run. We divided the fish identified
from the 125-locus genetic
stock identification data set of \citet{thompson2020complex}  as fall/late-fall into fall run and late-fall run according
to their sampling date at Keswick:  99 fish arriving before Dec 26 were classified as fall run and 89 fish arriving after Dec 26 were
classified as late-fall run.   The allele frequencies of the target SNPs at the remaining amplicons
were
estimated for each of these four groups: winter, spring, fall, and late-fall, and
amplicons showing large allele frequency differences between late-fall fish and all other runs were retained.

@@

Using the whole genome sequencing data of \citet{thompson2020complex}, we also sought
genomic regions that might provide diagnostic markers for the Sacramento winter-run
Chinook salmon.  This ecotype is highly differentiated from all other Chinook salmon populations
throughout its genome, which complicates the process of finding diverged genome regions using
a simple association study, with only 16 samples, as done to identify the LFAR.  Consequently we pursued a bespoke
analysis to find regions expected to have large allele frequency differences between
the winter-run and all other ecotypes of Chinook salmon in California's Central Valley.   This analysis
screened for regions with large allele frequency differences across large blocks of the genome, and is
described in the supplemental text section \ref{sec:wrap-methods}.

@@

We include six amplicons that capture the eight  SNPs within the ``Region of Strongest Assocation'' (RoSA)
between spring-run and fall-run fish, near the ROCK1 and GREB1L genes,  on Chromosome 28 listed in Table~S3 of \citet{thompson2020complex}.
We add three more amplicons in the same RoSA region to the panel, as well as one more amplicon that
allows genotyping of snp670329, which was discovered and described by \citet{thompson2019anthropogenic}.
Including all these markers in the panel allows them to be routinely typed, creating
 a database for understanding the spatial and temporal
distribution of the alleles in this genomic region; however these markers are omitted from likelihood calculations for
GSI, since their distribution across populations in the Feather River has been clearly and evidently
influenced by hatchery practices (see Results), it is not necessary to include them for accurate assignment,
and because the 12 SNPs within these three amplicons are typically in profound linkage disequilibrium (LD)
making them inappropriate for use as separate markers in most software for population assignment or GSI, that
typically assumes markers are not in LD within the component populations of the baseline.

@@

The baseline also includes three amplicons with genetic variation
in two genes, VGLL3 and SIX6,
found to be associated with phenotypes related to reproductive maturity in both
Atlantic \citep{barson2015sex} and Pacific \citep{waters2021heterogeneous} salmonids. They
provide additional variation for population genetic analyses and to
monitor populations for potential phenotypic patterns.

We include an assay from a novel locus that targets the sex-determining region ({\em sdY}) on the Y chromosome in Chinook salmon. We designed this assay to target the SNP described by \citet{bertho2022nonfunctional} that differentiates the functional {\em sdY} and a nonfunctional {\em sdY} pseudogene.  Examination of whole-genome sequence data in multiple populations of California Chinook salmon \citep{thompson2020complex} shows that this
marker characterizes genetic sex in diverse California salmon populations more accurately than previously published assays  \citep{von2015development}.
Finally, marker Ots\_coho001\_05\_32691399, which is fixed for alternate alleles in
coho salmon and Chinook salmon, is included to diagnose when samplers have unwittingly
collected coho salmon.

@@

The novel microhaplotype markers and the SNPlicon targets were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both of these
groups of markers we identified their locations within the genome by mapping the
consensus sequences (roughly 100--300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0, GenBank accession GCA\_002872995.1)
and version 2  (Otsh\_v2.0, GenBank accession GCA\_018296145.1) genome assemblies,
using both bwa mem \citep{bwa-mem2009} and the BLAST-like Alignment tool, BLAT \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers to
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

Amplicon sequencing was performed with the Genotyping-in-Thousands-by-sequencing (GT-seq)
method \citep{campbell2015genotyping}. All loci were
multiplexed with primer concentrations ranging from 0.083 uM to 0.5 uM to increase
uniformity of read depths across all loci. Sample normalization was
performed using the SequalPrep Normalization Plate Kit (Applied Biosystems) according
to the manufacturer's instructions. Samples were pooled by plate
and purified using Agencourt AMPure XP magnetic beads.
Libraries were quantified either by qPCR with the NEBNext Library Quant Kit for Illumina (New
England BioLabs) or with the Qubit dsDNA HS Assay Kit (Thermo Fisher Scientific)
before dilution and pooling for sequencing.
All samples were sequenced on a MiSeq using $2\times 75$-cycle paired-end sequencing protocols.
Primer sequences and concentrations for making the primer pool are in Supplemental Data~1.

@@

Basic data summaries were calculated for each population/collection
in the reference baseline
using simple operations in the R programming language, Version 4.3
\citep{rcore}. Summaries provided are:
\begin {itemize}
\item the average over loci of the number of fish
in which each locus was successfully genotyped, $\bar{N}$;
\item the average number of alleles per locus in each collection with all collections
downsampled to the smallest number of genotyped individuals at each
locus, $\bar{N}_\mathrm{A,ss}$;
\item the fraction of polymorphic
loci in each collection after subsampling to the smallest number of genotyped
fish, $\bar{P}_\mathrm{poly,ss}$;
\item the expected heterozygosity as the average over loci
of one minus the frequencies of homozygotes at each locus expected from the
estimated allele frequencies, $\bar{H}_\mathrm{exp}$; and
\item the observed heterozygosity as the average over
loci of the fraction of heterozygous genotypes, $\bar{H}_\mathrm{obs}$.
\end{itemize}
When subsampling each collection to the minimum number of genotyped
fish across all collections at a locus, individuals were subsampled without
replacement, taking the average of 1,000 different random subsamples.

We calculated \citet{weir1984estimating}'s pairwise $F_\mathrm{ST}$ between
all collections, using the {\footnotesize\tt pp.fst} function from the R
package `hierfstat' \citep{hierfstat}. Population
structure in the data was evaluated
using the program STRUCTURE \citep{pritchard2000inference,falush2003inference}.
For this analysis, we omitted the run-timing associated markers.
We performed 20 replicate runs at each value of the assumed number of
subpopulations, $K$, in $\{2, 3, 4, 5, 6, 7\}$
using default settings of the program.  STRUCTURE output was
summarized and visualized using CLUMPAK
\citep{kopelman2015clumpak}.
